---
title: "Relative Survival using the Pohar Perme estimator"
output: html_notebook
---

#advantages
*when cause specific mortality is not available or not good
*excess mortality in a group of patients
*epidemiological rather than RCT measure


####additive models 
*2 hazards - pop mortality + excess hazard due to disease = observed hazard
#hakulinen-tenkanen additive survival model
*k stratum for relevant predictor variables (age, sex, etc co-variates)
*life table estimated for each co-variate
*excess hazard is a multiplicative function of each covariates in each strata
*interval specific observed and expected survival proportions
*glm with binomial error strucutre with log-log link function / expected survival
#glm with poisson error structure
*same grouping as H-T
*excess hazard is a multiplicative function of each covariates in each strata
*deaths with a poisson distribution
*glm with poisson error structure and ln offset
#Esteve additive model
*doesn't do strata as those before
*individual data and estimates coeffecients using MLE approach


####multiplicative models
*2 hazards - pop mortality X excess hazard due to disease
*does NOT assume the observed hazard is greater than the pop hazard
#Andersen multiplicative model
*same as cox regression model with time dependent co-variate with coeffecient = 1.


####transformative models
*makes no assumption about the relationship between the observed, excess and pop hazard
*individual survival times are first transformed by cumulative distribution function of person of x age, y sex, z co-variates etc 
*distribution function is calculated from general population mortality data
*transforming to new scale takes population hazard into account leaving the disease specific hazard, which can be modelled using cox model for example. 


# Load the data
```{r}
library(relsurv)
library(survival)
library(splines)
library(date)

data(rdata) #to add the dataset
data(slopop) #to add the dataset
summary(slopop)
str(rdata)
```
# Fit the model and print the coefficients
*time - survival time variable
*cens - censoring variable
*ratetable - matches up the data in the pop and survival datasets
*int - no. of yearly intervals in the rsadd function (single number or vector)
*na.action - how to handle missing data points
*init - initial value
*control - specifies the number of iterations

*by default the youngest age group is the reference
*positive coeffecients mean that survival is better
*ratetable - array with all hazards per unit day
*see appendix in paper for more info on ratetables
#additive model with binomial error structure
``` {r}
attributes(slopop)$dimid #to match observed and pop data through function ratetable
f = rsadd(Surv(time,cens)~sex+as.factor(agegr)+ ratetable(age=age*365.24,sex=sex,year=year),
data=rdata,ratetable=slopop,int=5,method="glm.bin")
summary(f)
```
#additive model with poisson error structure
```{r}
rsadd(Surv(time,cens)~sex+as.factor(agegr)+
  ratetable(age=age*365.24,sex=sex,year=year),
data=rdata,ratetable=slopop,int=5,method="glm.poi") 
```
#
```{r}
rsadd(Surv(time,cens)~sex+as.factor(agegr)+
  ratetable(age=age*365.24,sex=sex,year=year),
data=rdata,ratetable=slopop,int=5) 
```
#multiplicative model
```{r}
rsmul(Surv(time,cens)~sex+as.factor(agegr)+
  ratetable(age=age*365.24,sex=sex,year=year),
data=rdata,ratetable=slopop,int=5) 
```
#transformative model
```{r}
rstrans(Surv(time,cens)~sex+as.factor(agegr)+ ratetable(age=age*365.24,sex=sex,year=year),
  data=rdata,ratetable=slopop,int=5)
```


